JD notes on effects

### 2022-07-24 manuscript/body.tex
A little rest!

When dealing with nonlinear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- as illustrated by the two examples above. In such cases, bias correction is needed when back-transforming the estimates to the original scales. The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}. Here, we describe and implement a different approach -- observed-value-based approach for bias correction. Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only. Our formulation is more general and can easily be extended to other link functions other than logistic.

@@

An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}. The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}. More specifically:
\begin{itemize}
\item compute linear predictor associated with whole sample of the non-focal predictors, $\etanfj = \sum \uset{\hat{\beta}} \uset{x}$
\item compute linear predictor associated with the chosen values focal predictor, $\etahfi = \hat{\beta}^f x^f$
\item for every value of the focal linear predictor, $\etahfi$, compute
%
\begin{align}\label{eq:pop_eta} 
\etahfj(\etahfi, \etanfj)  &= \etahfi + \etanfj \nonumber \\
&= \etahfj(x^f, \uset{x}).
\end{align}
\end{itemize}
%

@@

In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging. In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale. In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results. However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population. In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the observed-value-based approach summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This might not be a problem in the observed-value-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.

@@

We simulated data motivated by the water, sanitation, and hygiene (WaSH) study in which we were interested in investing the contribution of demographic and socio-economic factors to improved WaSH indicators among slum dwellers in Nairobi, Kenya. We used the mean-based approach to generate the predicted probabilities. However, we noticed that the predictions consistently over- or under- estimated the observed proportions; and did not align well with the observed data points. To demonstrate this, we consider a binary-outcome simulation with two input variables (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and observed-value-based central estimates.} The \emph{truth} is based on the true simulation parameters. If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A. In the case of strong effect size, however, the two approaches produce very close estimates -- B. In both cases, the observed-value-based estimates are closer to the truth. The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model. Mean-based approach suffers from nonlinear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal predictor(s) with the strong effect. However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small. Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned. The horizontal dashed lines correspond to the respective averages. The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin.} 
\label{fig:pred_bin_plots}
\end{figure}

If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B. One possible reason for the variations we see in \FREF{pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.

@@

\jd{We don't \emph{need} more than one model; better to say there is more than one question we could ask. In this case, we should make a distinction between direct and total effects.}
To make inferences about this mediated effect, we need more than one model. First, the regression of $z$ on $x$, i.e., the \emph{non-mediated} (or univariate) model which tells on the \emph{total} effect of $x$ on $z$; and second, the regression of $z$ on $x$ and $y$, i.e., the \emph{mediated} (or multivariate or full) model. We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated. In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.

\jd{“no” direct effect is a weird and confusing choice. The absence of direct effect is not what's causing the bias: it's the presence of the indirect effect. Please re-do with a relatively small (but non-zero) direct effect.}

\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and observed-value-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and observed-value-based approaches central estimates are identical since there is no any other additional (non-focal predictor) sources of bias. Consequently, in both approaches, the average estimated probability is very close to the observed proportion. B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the central estimates do not align with the observations. However, the observed-value-based approach still closely approximates the average proportion of the simulated data. The horizontal dashed lines are the respective average estimates and observed proportions. The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

\jd{I'd like to see a better explanation here as well. In my opinion, the reason for the mismatch is non-linear averaging over the non-focal variable; I don't think dilution is a good metaphor (a linear model could be equally diluted, but would not have this effect, I think?).}
From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. This is the total effect of $x$, which only tells us the influence of $x$ on $z$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our estimates do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).

\section{Discussion}

\jd{This doesn't seem like a good starting point. I guess the point is that you're worried about taking out random effects? Don't worry, and let's just take them all the way out!}
Our simulations examples focused on simple linear and logistic models due their wide range of usage and application. These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. The logic for extending to more complex models, including other forms of nonlinear link functions is very straightforward. In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes. In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.

\jd{and now we're left with only one, boring Discussion \P. Maybe merge Discussion and Conclusion.}
Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the observed-value-based approach has a potential of yielding results which are more consistent with the observed data. We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.

\section{Conclusion}

Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest. However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity and complexities in interpretation.

The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models. For example, effect plots could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.

From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

>>>>

When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the non-linear transformation of the response variable -- as illustrated by the two examples above. In such cases, bias correction is needed when back-transforming the estimates to the original scales. The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}. Here, we describe and implement a different approach -- observed-value-based approach for bias correction. Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only. Our formulation is more general and can easily be extended to other link functions other than logistic.

@@

An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}. The non-linear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}. More specifically:
\begin{itemize}
\item compute linear predictor associated with whole sample of the non-focal predictors, $\etanfj = \sum \uset{\hat{\beta}} \uset{x}$
\item compute linear predictor associated with the chosen values focal predictor, $\etahfi = \hat{\beta}^f x^f$
\item for every value of the focal linear predictor, $\etahfi$, compute
%
\begin{align}\label{eq:pop_eta} 
\etahfj(\etahfi, \etanfj)  &= \etahfi + \etanfj \nonumber \\
&= \etahfj(x^f, \uset{x}).
\end{align}
\end{itemize}
%

@@

In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the non-linear averaging. In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale. In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results. However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population. In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the observed-value-based approach summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This might not be a problem in the observed-value-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.

@@

We simulated data motivated by the water, sanitation, and hygiene (WaSH) study in which we were interested in investing the contribution of demographic and socio-economic factors to improved WaSH indicators among slum dwellers in Nairobi, Kenya. We used the mean-based approach to generate the predicted probabilities. However, we noticed that the predictions consistently over- or under- estimated the observed proportions; and did not align well with the observed data points. To demonstrate this, we consider a binary-outcome simulation with two input variables (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and observed-value-based central estimates.} The \emph{truth} is based on the true simulation parameters. If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A. In the case of strong effect size, however, the two approaches produce very close estimates -- B. In both cases, the observed-value-based estimates are closer to the truth. The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear link function in the logistic model. Mean-based approach suffers from non-linear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of non-linear averaging is driven by the non-focal predictor(s) with the strong effect. However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small. Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the non-linear averaging for the reasons already mentioned. The horizontal dashed lines correspond to the respective averages. The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin.} 
\label{fig:pred_bin_plots}
\end{figure}

If there was no effect of non-linear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B. One possible reason for the variations we see in \FREF{pred_bin_plots} is the non-linear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.

@@

To make inferences about this mediated effect, we can ask questions about the \emph{total} and \emph{direct} effects. In the first case, we want to fit a univariate model which regresses $z$ on $x$, excluding $y$. We refer to this as \emph{non-mediated} model. In the second case, we want to fit a multivariate model which regresses $z$ on $x$ and $y$. We refer to this as \emph{mediated} model. We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated. In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and observed-value-based estimates for non-mediated (total) effect models and mediated (direct).} A: In the absence of mediator variable, the central estimate aligns well with the observed data. B: In the presence of indirect effect (mediator variable included), the central estimates do not align with the observations. Since the models in A and B are both linear models with identity link function, both approaches (mean-based and observed-value) give the same estimates. The horizontal dashed lines are the respective average estimates and observed means ($\bar{\hat{z}} \approx \bar{z}$). The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}
%
From \FREF{pred_mediated_plots}A, we see what we would expect if we simulated a univariate model with no non-focal predictors, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. This is the total effect of $x$, which only tells us the influence of $x$ on $z$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and match the observed values using both approaches. However, if we control for the mediator variable $y$, the estimates do not necessarily match the observed values since $x$ has indirect effect on $z$. By controlling for the mediator variable, we able to get an indication of the direct (vary week) effect of $x$ on $z$ (see \FREF{pred_mediated_plots}B).

\section{Discussion and conclusion}

Generalized linear (mixed) models have a wide usage in various fields including public health. In a model involving difficult-to-interpret coefficient estimates, an outcome plot can aid in understanding and summarizing the results. In particular, if the goal is to capture every uncertainty in the model for a particular focal predictor or if we are interested on total effect, then a prediction plot would be appropriate. On the hand, an effect plot is preferable if we want to focus on the uncertainty associated with a focal predictor only or if we are interested in the direct effect.

Mean-based approach is widely to generate outcome plots. However, in a model with complex interaction, MPVs or categorical variables, it is sensitive to the choice of the reference point. We have demonstrated that non-focal model-center-based reference point is generally a stable choice and provide estimates which are more consistent with the observed quantities.

In a model with non-linear link function such as a logistic or exponential function, the generated central estimate curve may not match well with the observed data, i.e., our description for ``bias''. In such such a model, observed-value-based approach provides a way to generate more consistent estimates and is preferable over conventional mean-based approach.

The argument and results we present in this paper support a greater need for a shift on focus on how to summarize regression models. From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

Our simulations examples focused on simple linear and logistic models due their wide range of usage and application. These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. The logic for extending to more complex models, including other forms of non-linear link functions is very straightforward. In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes. In addition, our \proglang{R} package implementation already extends to and supports most of the non-linear link functions and mixed model framework, including multivariate binary outcome models.
